Multireference Alignment Notes

### 2014-01-29 Amit_notes/Procrustes.tex
Updated notes from Amit

@@

Assume we have a signal that consists of $n$ features, each which lie in $d$-dimensional space (e.g. if we have a 10-atom molecule in three-dimensional space, then $n=10$ and $d=3$).

Denote the features (atoms) as $\x_1, \dots, x_n \in \mathbb{R}^d$.

Assume that we observe $m$ rotated noisy copies of the features
$$y_i^{(k)} = O_k x_i + \xi_^{(k)}, k=1, \dots, m, i=1, \dots, n$$

We would like to construct the original features $x_1, \dots, x_n$ from the rotated noisy copies $y_i^{(k)}$

Note that, up to noise
$$ \langle y_i^{(k)}, y_j^{(k)} \rangle  \approx \langle x_i, x_j \rangle $$
because the rotations preserve inner products.

Therefore,  we can approximate $\langle x_i, x_j \rangle$ as 
$$ \langle x_i, x_j \rangle  \approx \frac{1}{m} \sum_{k=1}^m \langle y_i^{(k)}, y_j^{(k)} \rangle $$

We can therefore estimate the matrix $X$, where $X_{ij} = \langle x_i, x_j \rangle$.

However, this is precisely the covariance matrix of the data $x_1, \dots, x_n$.

We can therefore recover the original signal using PCA/eigendecomposition.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}

\title{Solving the Procrustes Problem with PCA}

@@

\maketitle

Assume we have a signal that consists of $n$ features, each of which lie in $d$-dimensional space (e.g. if we have a 10-atom molecule in three-dimensional space, then $n=10$ and $d=3$).
%
Denote the features as $x_1, \dots, x_n \in \mathbb{R}^d$.

We observe $m$ rotated noisy copies of the features
\begin{equation}
y_i^{(k)} = O_k x_i + \xi_i^{(k)}, k=1, \dots, m, i=1, \dots, n,
\end{equation}
where $O_k$ is an orthoganol transform, and $\xi_i^{(k)}$ is Gaussian white noise. 
%
We would like to estimate the original signal $x_1, \dots, x_n$ from the rotated noisy copies $y_i^{(k)}$.

Note that, because orthoganol transforms preserve inner products,
\begin{equation}
\langle y_i^{(k)}, y_j^{(k)} \rangle  \approx \langle x_i, x_j \rangle 
\end{equation}
(up to noise).

Therefore,  we can estimate $\langle x_i, x_j \rangle$ as 
\begin{equation} \label{eq:ip_est}
\langle x_i, x_j \rangle  \approx \frac{1}{m} \sum_{k=1}^m \langle y_i^{(k)}, y_j^{(k)} \rangle 
\end{equation}

We then construct the matrix $C$, where $C_{ij} = \langle x_i, x_j \rangle$, using the estimates in \eqref{eq:ip_est}.
%
However, $C$ is precisely the covariance matrix of the data $x_1, \dots, x_n$. 
%
We can therefore recover $x_1, \dots, x_n$ using eigendecomposition:
\begin{equation}
\begin{bmatrix}
x_1^T \\
\vdots \\
x_n^T 
\end{bmatrix} = 
\begin{bmatrix}
| & | &  & | \\
\phi_1 & \phi_2 & \cdots & \phi_d \\
| & | &  & | \\
\end{bmatrix}
\begin{bmatrix}
\sqrt{\lambda_1} & 0 & 0 & \cdots & 0 \\
0 & \sqrt{\lambda_2} & 0 & \cdots & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
0 & 0 & 0 & \cdots & \sqrt{\lambda_d} 
\end{bmatrix}
\end{equation}
where $\lambda_1, \dots, \lambda_n$ and $\phi_1, \dots, \phi_n$ are the eigenvalues and eigenvectors, respectively, of $C$, and they are ordered such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_n$.

### 2014-01-29 Amit_notes/multireference_alignment.tex
Updated notes from Amit

\documentclass[12pt]{article}

@@

We have images $I_1, \dots, I_n$

Want to find rotations $R_1, \dots, R_n \in SO(2)$ such that the images are optimally aligned

Define a cost function $c(R_i I_i, R_j I_j)$.
%
Then we can write the objective function 
$$\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(R_i I_i, R_j I_j)$$

Since the cost is invariant to global rotations, we can rewrite it as
$$\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(I_i, R_i^T R_j I_j)$$

We can discretize the space of rotations into $L$ intervals, and write each rotation as a number $l \in \mathbb{Z}_L$. 
%
We also define a cost function between a {\em specific} pair of images $I_i$ and $I_j$ as $c_{ij}$. 
%
This cost function is only a function of the difference in rotations, $l_i - l_j$. 
%
Therefore, we can rewrite the optimization problem as
$$\min_{l_1, \dots, l_n \in \mathbb{Z}_L} \sum_{i \ne j} c_{ij} (l_i - l_j)$$

We now note that each rotation $l_i-l_j \in \mathbb{Z}_L$ maps to a {\em cyclic} permutation matrix $\pi_{ij} \in \mathbb{R}^{L \times L}$, because each rotation can be thought of as a cyclic permutation of the $L$ ``ticks'' on the unit circle.
%
However, the set of cyclic permutation matrices is non-convex (e.g. because the average of two permutations matrices is not a permutation matrix). 
%
The convex hull of permutation matrices is the set of all doubly stochastic matrices:
$$\pi_{ij}(k,l) \ge 0  \forall k, l $$
$$\pi_{ij} \mathbb{1} = \mathbb{1}$$
$$\mathbb{1}^T \pi_{ij} = \mathbb{1}^T$$

We also require the cyclic condition to hold:
$$\pi_{ij}(k,l) = \pi_{ij}(k+1, l+1)$$

We would also like {\em consistency} relations to hold, e.g. $\pi_{ij} \pi_{jk} = \pi_{ik}$.
%
This is trivially satisfied if $\exists \pi_1, \dots, \pi_n$ such that $\pi_{ij} = \pi_i^T \pi_j \forall i, j$. 
%
Therefore, we want
$$\Pi = \left[ \begin{array}{c} \pi_1^T \\ \dots \\ \pi_n^T \end{array} \right] [\pi_1 \dots \pi_n]$$
where $\Pi \in \mathbb{R}^{nL \times nL}$ is the Gram matrix, where the $i,j$ block of $\Pi$ is $\pi_{ij} \in \mathbb{R}^{L \times L}$.

This is true if $rank(\Pi) = L$ and $\Pi$ is positive semidefinite. 
%
We relax the constrant that $rank(\Pi) = L$, and only require that $\Pi$ is psd.

We also require that
$$\pi_{ii} = I){L \times L}$$

We then compute the cost function $c_{ij}$ for every possible $l_i - l_j \in \mathbb{Z}_L$, and put these values into the matrix $C_{ij} \in \mathbb{R}^{L \times L}$.
%
The objective function corresponding to a particular permutation matrix can then be written as 
$$\sum_{i \ne j} Tr(C_{ij} \pi_{ij})$$

Or, if we write it in full matrix notation
$$\min_{\Pi} Tr(C^T \Pi)$$

>>>>

\documentclass[12pt]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{dsfont}

\title{Multireference Alignment Notes}

@@

\maketitle

We are given images $I_1, \dots, I_n$.

We want to find rotations $R_1, \dots, R_n \in SO(2)$ such that the images are optimally aligned.

We define a cost function $c(R_i I_i, R_j I_j)$.
%
We would like to solve the optimization problem
\begin{equation} \label{eq:opt1}
\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(R_i I_i, R_j I_j)
\end{equation}

Since the cost is invariant to global rotations, we can rewrite \eqref{eq:opt1} as
\begin{equation}
\min_{R_1, \dots R_n \in SO(2)} \sum_{i \ne j} c(I_i, R_i^T R_j I_j)
\end{equation}

We discretize the space of rotations into $L$ intervals, and write each rotation as a number $l_i \in \mathbb{Z}_L, i=1, \dots, n$. 
%
We also define a cost function between a {\em specific} pair of images $I_i$ and $I_j$ as $c_{ij}(l_i - l_j)$; 
this cost function is only a function of the difference in rotations, $l_i - l_j$. 
%
we rewrite the optimization problem as
$$\min_{l_1, \dots, l_n \in \mathbb{Z}_L} \sum_{i \ne j} c_{ij} (l_i - l_j)$$

We now note that each rotation $l_i-l_j \in \mathbb{Z}_L$ maps to a {\em cyclic} permutation matrix $\pi_{ij} \in \mathbb{R}^{L \times L}$, because each rotation can be thought of as a cyclic permutation of the $L$ ``ticks'' on the unit circle.
%
We will optimize over the space of cyclic permutation matrices, rather than the space of rotation matrices.

However, the set of cyclic permutation matrices is non-convex (e.g. because the average of two permutations matrices is not a permutation matrix). 
%
The convex hull of permutation matrices can be shown to be the set of all doubly stochastic matrices, which satisfy the following constraints
\begin{equation} \label{eq:doublestoch}
\begin{aligned}
\pi_{ij}(k,l) &  \ge 0 \forall k, l \\
\pi_{ij} \mathds{1} & = \mathds{1}\\
\mathds{1}^T \pi_{ij} & = \mathds{1}^T 
\end{aligned}
\end{equation}

We also require the cyclic condition to hold:
\begin{equation} \label{eq:cyclic_const}
\pi_{ij}(k,l) = \pi_{ij}(k+1, l+1)
\end{equation}

We would also like {\em consistency} relations to hold, e.g. $\pi_{ij} \pi_{jk} = \pi_{ik}$.
%
This is trivially satisfied if $\exists \pi_1, \dots, \pi_n$ such that $\pi_{ij} = \pi_i^T \pi_j \forall i, j$. 
%
Therefore, we want
\begin{equation} \label{eq:Gram_const}
\Pi = 
\begin{bmatrix}
\pi_{11} & \pi_{12} & \cdots & \pi_{1n} \\
\pi_{21} & \pi_{22} & \cdots  & \pi_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
\pi_{n1} & \pi_{n2} & \cdots  & \pi_{nn} 
\end{bmatrix} = 
\begin{bmatrix}
\pi_1^T \\
\vdots \\
\pi_n^T 
\end{bmatrix}
\begin{bmatrix}
\pi_1^T \cdots \pi_n^T 
\end{bmatrix}
\end{equation}
where $\Pi \in \mathbb{R}^{nL \times nL}$ is the Gram matrix, where the $i,j$ block of $\Pi$ is $\pi_{ij} \in \mathbb{R}^{L \times L}$.

\eqref{eq:Gram_const} holds if $rank(\Pi) = L$ and $\Pi$ is positive semidefinite. 
%
We relax the constrant that $rank(\Pi) = L$, and only require  
\begin{equation} \label{eq:psd_const}
\Pi \succeq 0
\end{equation}

We also require that
\begin{equation} \label{eq:identity_const}
\pi_{ii} = I_{L \times L}
\end{equation}

We then construct the matrices $C_{ij}  \in \mathbb{R}^{L \times L}$ for $i, j= 1, \dots, n$, where
\begin{equation}
C_{ij}(l_1, l_2) = c_{ij}(l_1 - l_2)
\end{equation}

The objective function can then be written as
\begin{equation} \label{eq:one_cost}
\min_{\pi_{ij}} \sum_{i \ne j} Tr(C_{ij} \pi_{ij})
\end{equation}

We then define the matrix $C \in \mathbb{R}^{nL \times nL}$, where
$$C = \begin{bmatrix}
 C_{11} & C_{12} & \cdots & C_{1n} \\
 C_{21} & C_{22} & \cdots & C_{2n} \\
\vdots & \vdots & \ddots & \vdots \\ 
C_{n1} & C_{n2} & \cdots & C_{nn} \\
\end{bmatrix} $$

We can then write \eqref{eq:one_cost} as 
\begin{equation} \label{eq:total_cost}
\min_{\Pi} Tr(C^T \Pi)
\end{equation}

\eqref{eq:total_cost}, togther with constraints \eqref{eq:doublestoch}, \eqref{eq:cyclic_const}, \eqref{eq:psd_const}, and \eqref{eq:identity_const}, define the semidefinite program that we would like to solve.
